```{r message=FALSE, warning=FALSE, include=FALSE}
library(flextable)
```

### Activity 1.1 {-}

25 participants were enrolled in a 3-week weight loss program. The following data present the weight loss (in grams) of the participants.

```{r act-1-1, echo=FALSE}
#| tbl-cap: Weight loss (g) for 25 participants

tab1_1 <- tibble::tribble(
    ~v1, ~v2, ~v3, ~v4, ~v5,
    255L, 198L, 283L, 312L, 283L,
    57L, 85L, 312L, 142L, 113L,
    227L, 283L, 255L, 340L, 142L,
    113L, 312L, 227L, 85L, 170L,
    255L, 198L, 113L, 227L, 255L
)

flextable(tab1_1) |>
    delete_part(part = "header") |>
    theme_booktabs()
```

a) These data have been saved as `Activity_1.1.rds`. Read the data into your software package.
b) What type of data are these?
c) Construct an appropriate graph to display the distribution of participants' weight loss. Provide appropriate labels for the axes and give the graph an appropriate title.

### Activity 1.2 {-}
Which of the following statements are true? The more dispersed, or spread out, a set of observations are:

a) The smaller the mean value
b) The larger the standard deviation
c) The smaller the variance


### Activity 1.3 {-}
Estimate the mean, median, standard deviation, range and interquartile range for the data  `Activity_1.3.rds`, available on Moodle.

### Activity 1.4 {-}
Data of diastolic blood pressure (BP) of a sample of study participants are provided in the datasets `Activity_1.4.rds`. Compute the mean, median, range and SD of diastolic BP.

### Activity 1.5 {-}
The ages of 100 study participants have been saved as `Activity_1.5.rds`. Estimate the:

a. mean and median;
b. standard deviation and interquartile range;
c. range.

Plot the data using a histogram and boxpolot. Is there anything unusual about the ages? What do you think is a possible explanation for this?

A clean version of the data have been saved as `Activity_1.5_clean.rds`. Recalculate the summary statistics and recreate the plots using the clean data.

Based on this exercise, what is your advice on coding unusual or missing values in data?
