```{r message=FALSE, warning=FALSE, include=FALSE}
library(jmv)
```

We will demonstrate using R for correlation and simple linear regression using the dataset `mod08_lung_function.csv`.

```{r}
lung <- read.csv("data/examples/mod08_lung_function.csv")
```

## Creating a scatter plot

We can use the `plot` function to create a scatter plot to explore the association between height and FVC, assigning meaningful labels with the `xlab` and `ylab` commands:

```{r}
plot(x=lung$Height, y=lung$FVC, 
     xlab="Height (cm)", 
     ylab="Forced vital capacity (L)")
```

To add a fitted line, we can use the `abline()` function which adds a straight line to the plot. The equation of this straight line will be determined from the estimated regression line, which we specify with the `lm()` function, which fits a *linear model*.

The basic syntax of the `lm()` function is: `lm(y ~ x)` where `y` represents the *outcome* variable, and `x` represents the *explanatory* variable. Putting this all together:

```{r}
plot(x=lung$Height, y=lung$FVC,
     xlab="Height (cm)",
     ylab="Forced vital capacity (L)")

abline(lm(lung$FVC ~ lung$Height))
```

Note: to obtain output similar to Jamovi output, we can use the `scatr` package:

```{r}
# Install scatr if required:
# install.package("scatr")
library(scatr)

scat(data=lung, x="Height", y="FVC", line="linear")
```


## Calculating a correlation coefficient {-}

We can use the `corrMatrix` function in the Jamovi package to calculate a Pearson's correlation coefficient:

```{r}
corrMatrix(data=lung, vars=c(Height, FVC))
```

## Fitting a simple linear regression model

We can use the `lm` function to fit a simple linear regression model, specifying the model as `y ~ x` where `y` represents the *outcome* variable, and `x` represents the *explanatory* variable. Using `mod08_lung_function.csv`, we can quantify the relationship between FVC and height:

```{r}
lm(FVC ~ Height, data=lung)
```

The default output from the `lm` function is rather sparse. We can obtain much more useful information by defining the linear regression model as an object, then using the `summary()` function:

```{r}
model <- lm(FVC ~ Height, data=lung)
summary(model)
```

Finally, we can obtain 95% confidence intervals for the regression coefficients using the `confint` function:

```{r}
confint(model)
```

Note that the output for R looks slightly different from the Jamovi output, the numerical values are identical.

## Plotting residuals from a simple linear regression

We can use the `resid` function to obtain the residuals from a saved model. These residuals can then be plotted using a density plot in the usual way:

```{r}
residuals <- resid(model)
plot(density(residuals))
```
