---
title: "surveymv for jamovi"
---

An alternative way of producing barcharts is via a package called `surveymv`. Unfortunately, surveymv is not hosted on the standard package repository, we need to install the package from github.com. This is a straight-forward process:

```{r}
#| eval: false
#| include: true

install.packages("devtools")
library(devtools)
install_github("raviselker/surveymv")
```

Once we install `surveymv`, we load it using the standard `library()` command:

```{r}
#| eval: false
#| include: true

library(surveymv)
```

`surveymv` has only one function: `surveyPlot`, with the following syntax:

```
surveyPlot(
  data = babies,
  vars = "Birth weight",
  group = "Age group",
  type = "stacked",
  freq = "perc")
```

We specify our data (`data=`), and the main variable to be plotted (`vars=`). If we have a grouping variable, we specify a `group=` variable. We define the chart to be either a stacked (`type = "stacked"`) or clustered (`type = "grouped"`) bar chart, and specify whether to plot frequencies (`freq = "count"`) or percentages (`freq = "perc"`).

To demonstrate, we will recreate @fig-bar-4, a figure of the relative frequencies of sex within stage of disease:

```{r message=FALSE}
library(surveymv)

pbc <- readRDS("data/activities/mod01_pbc.rds")

pbc$sex <- factor(pbc$sex,
                  levels = c(1, 2),
                  labels = c("Male", "Female"))
pbc$stage <- factor(pbc$stage,
                    levels = c(1, 2, 3, 4),
                    labels = c("Stage 1", "Stage 2", "Stage 3", "Stage 4"))

surveyPlot(
  data = pbc,
  vars = "sex",
  group = "stage",
  type = "stacked",
  freq = "perc")
```

While this produces a bar chart with horizontal bars, it often performs better with labelling of the groups than the previous method.